---
title: "confined_bulkRNAseq_analysis"
author: "Miranda Hunter"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries:
``` {r, message = F, warning = F}
library(DESeq2)
library(tidyverse)
library(scales)
library(pals)
library(cowplot)
library(fgsea)
library(msigdbr)
library(data.table)
library(mvhfunctions)
```

## Import data and run DESeq2.

Import counts matrices and organize data.
``` {r}
wd <- '/Users/hunterm/Library/CloudStorage/GoogleDrive-hunterm@mskcc.org/.shortcut-targets-by-id/1L5uwxnNO44Pd7GBMBhMGO9vvsi7uifYZ/White Lab Gsuite Main Drive/Lab members/Miranda Hunter_GSuite/Miranda_R_new/bulk_rnaseq_datasets/confined_A375/' 
setwd(wd)

groups <- c("NC1", "NC2", "NC3", "C1", "C2", "C3") 

# import counts tables
for (group in groups) {
  path <- paste0(wd, group, "_ReadsPerGene.out.tab") # create a path saying where each counts matrix file is stored
  data <- read.delim(file = path, skip = 3) # read in the counts matrix file
  data[3:4] <- NULL # the last two columns of the matrix don't have useful info so just remove them
  colnames(data)[1:2] <- c("Ensembl", group) # rename columns to have better names
  assign(group,data) # store counts matrix for each group as the group name
  rm(data)
}

# create counts matrix containing all samples
counts_matrix <- merge(x = eval(parse(text = groups[1])),
                       y = eval(parse(text = groups[2])),
                       by = "Ensembl")

for (ii in 3:length(groups)) {
  group <- as.character(groups[ii])
  counts_matrix <- merge(x = counts_matrix,
                         y = eval(parse(text = group)),
                         by = "Ensembl")
}

counts_matrix <- counts_matrix %>% column_to_rownames(var = "Ensembl")

rm('C1', 'C2', 'C3', 'NC1', 'NC2', 'NC3')
```



Create DESeq2 dataset.
``` {r}
# create dataframe with your groups and the number of each replicate.
col_data <- data.frame(condition = c("NC", "NC", "NC", "C", "C", "C"), replicate = rep(c("1", "2", "3"), 2))
rownames(col_data) <- colnames(counts_matrix)
col_data$condition <- factor(col_data$condition)
col_data$replicate <- factor(col_data$replicate)
col_data$condition <- relevel(col_data$condition, ref = "NC")

deseq_matrix <- DESeqDataSetFromMatrix(countData = counts_matrix,
                                       colData = col_data,
                                       design = ~condition)
deseq_matrix
```

Pre filter low count genes:
``` {r}
keep <- rowSums(counts(deseq_matrix)) >= 1 # remove any gene with 1 or fewer counts 
deseq_matrix <- deseq_matrix[keep, ]
```

Run DESeq2.
``` {r}
deseq_matrix <- DESeq(deseq_matrix, betaPrior = T)
resultsNames(deseq_matrix)
```

Make some comparisons across groups.
``` {r}
# for contrast: whatever the fold change is relative to goes last!!!!
dp_res <- results(deseq_matrix, contrast = c("condition", "C", "NC"))

summary(dp_res)
```

Normalization and PCA.
``` {r}
deseq_matrix_norm <- vst(deseq_matrix, blind = F)

# PCA
pcaData <- plotPCA(object = deseq_matrix_norm,
                   intgroup = "condition",
                   returnData = F)
percentVar <- round(100 * attr(pcaData, "percentVar")) # calculate percent variance for the different PCs

pca_data <- plotPCA(object = deseq_matrix_norm,
                    intgroup = 'condition',
                    returnData = T)

ggplot(pca_data, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 4) +
  theme_classic() +
  xlab('PC1 (64% variance)') +
  ylab('PC2 (17% variance)') +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_line(size = 0.75),
        axis.title = element_text(size = 16, color = 'black'))
```


To get PC1 and PC2 genes: repeat PCA manually * with named genes from below *
``` {r}
vst_norm_matrix <- assay(deseq_matrix_norm) # extract normalized counts matrix from deseq_matrix_norm calculated above

pca_data_for_extracting_genes <- prcomp(t(vst_norm_matrix)) # re-run PCA to extract genes

# extract data with the contribution of each gene to the PCA calculation:
pca_gene_info <- pca_data_for_extracting_genes$rotation %>%
  data.frame() %>%
  rownames_to_column(var = "gene")

n_genes <- 5 # select number of genes you want to look at

# PC1:
PC1_high <- pca_gene_info %>% arrange(-PC1) %>% head(., n_genes) %>% pull(gene) # pull out the PC1 high genes
PC1_low <- pca_gene_info %>% arrange(-PC1) %>% tail(., n_genes) %>% pull(gene) # pull out the PC1 low genes

# PC2:
PC2_high <- pca_gene_info %>% arrange(-PC2) %>% head(., n_genes) %>% pull(gene) # pull out the PC2 high genes
PC2_low <- pca_gene_info %>% arrange(-PC2) %>% tail(., n_genes) %>% pull(gene) # pull out the PC2 low genes

# print results
message('PC1 high genes:')
message(paste(PC1_high, sep = " ", collapse = ", "))
message('PC1 low genes:')
message(paste(PC1_low, sep = " ", collapse = ", "))
message('PC2 high genes:')
message(paste(PC2_high, sep = " ", collapse = ", "))
message('PC2 low genes:')
message(paste(PC2_low, sep = " ", collapse = ", "))

```

## Add gene names

Load gene list:
``` {r}
dp_res_df <- data.frame(dp_res) %>% rownames_to_column(var = "Ensembl")

```

Load conversion table:
``` {r}
library(biomaRt)
ensembl <- useEnsembl(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", mirror = "useast")
```


Extract gene symbols and ENSEMBL IDs:
``` {r}
data <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", 'description'),
              filters = "ensembl_gene_id",
              values = dp_res_df$Ensembl,
              mart = ensembl)
```

Add to DESeq2 table:
``` {r}
dp_res_df_named <- merge(x = dp_res_df,
                         y = data,
                         by.x = "Ensembl",
                         by.y = "ensembl_gene_id") %>% dplyr::select(Ensembl, hgnc_symbol, description, everything()) %>% arrange(-log2FoldChange)
```


Also add to counts matrix and normalized matrix:
``` {r}
counts_matrix <- counts_matrix %>% rownames_to_column(var = "Ensembl")
counts_matrix_named <- merge(x = counts_matrix,
                             y = data,
                             by.x = "Ensembl",
                             by.y = "ensembl_gene_id") %>% dplyr::select(Ensembl, hgnc_symbol, everything())

vst_norm_matrix <- data.frame(vst_norm_matrix) %>% rownames_to_column(var = "Ensembl")
vst_norm_matrix_named <- merge(x = vst_norm_matrix,
                         y = data,
                         by.x = "Ensembl",
                         by.y = "ensembl_gene_id") %>% dplyr::select(Ensembl, hgnc_symbol, everything())

```


Add ENSEMBL ID to hgnc_genes column in cases where there isn't a gene name. 
``` {r}
dp_res_df_named$hgnc_symbol <- case_when(
  dp_res_df_named$hgnc_symbol == "" ~ dp_res_df_named$Ensembl,
  TRUE ~ dp_res_df_named$hgnc_symbol
)

vst_norm_matrix_named$hgnc_symbol <- case_when(
  vst_norm_matrix_named$hgnc_symbol == "" ~ vst_norm_matrix_named$Ensembl,
  TRUE ~ vst_norm_matrix_named$hgnc_symbol
)

counts_matrix_named$hgnc_symbol <- case_when(
  counts_matrix_named$hgnc_symbol == "" ~ counts_matrix_named$Ensembl,
  TRUE ~ counts_matrix_named$hgnc_symbol
)
```


## Start here if loading in previous results.
``` {r}
load('/Users/hunterm/Library/CloudStorage/GoogleDrive-hunterm@mskcc.org/.shortcut-targets-by-id/1L5uwxnNO44Pd7GBMBhMGO9vvsi7uifYZ/White Lab Gsuite Main Drive/Lab members/Miranda Hunter_GSuite/Miranda_R_new/bulk_rnaseq_datasets/confined_A375/counts_matrix_named_confined.R')
load('/Users/hunterm/Library/CloudStorage/GoogleDrive-hunterm@mskcc.org/.shortcut-targets-by-id/1L5uwxnNO44Pd7GBMBhMGO9vvsi7uifYZ/White Lab Gsuite Main Drive/Lab members/Miranda Hunter_GSuite/Miranda_R_new/bulk_rnaseq_datasets/confined_A375/dp_res_df_named_confined.R')
load('/Users/hunterm/Library/CloudStorage/GoogleDrive-hunterm@mskcc.org/.shortcut-targets-by-id/1L5uwxnNO44Pd7GBMBhMGO9vvsi7uifYZ/White Lab Gsuite Main Drive/Lab members/Miranda Hunter_GSuite/Miranda_R_new/bulk_rnaseq_datasets/confined_A375/vst_norm_matrix_named_confined.R')
```


## Run GSEA

Load pathways.
``` {r}
human.genes <- msigdbr(species = "Homo sapiens")
pathways.GOBP <- filter(human.genes, gs_subcat == "GO:BP") %>% split(x = .$gene_symbol, f = .$gs_name) # GO BP
```


GSEA with GO biological processes pathway set.
``` {r}
gsea_genelist <- dp_res_df_named_confined %>% dplyr::select(hgnc_symbol, log2FoldChange) %>% deframe()
cutoff <- 0.25
gsea_genelist_filt <- gsea_genelist[gsea_genelist > cutoff | gsea_genelist < -cutoff]

set.seed(101) 
gsea_gobp_filt <- fgseaSimple(pathways = pathways.GOBP,
                              stats = gsea_genelist_filt,
                              nperm = 10000) %>%
  mutate(log10pval = -log10(pval)) %>%
  dplyr::select(pathway, pval, log10pval, everything()) %>%
  arrange(desc(NES)) 

gsea_gobp_filt
```




