---
title: "confined_bulkRNAseq_analysis"
author: "Miranda Hunter"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries:
``` {r, message = F, warning = F}
library(DESeq2)
library(tidyverse)
library(scales)
library(pals)
library(cowplot)
library(fgsea)
library(msigdbr)
library(data.table)
library(mvhfunctions)
```

## Import data and run DESeq2

# Import counts matrices and organize data.
``` {r}
wd <- '/Users/hunterm/Library/CloudStorage/GoogleDrive-hunterm@mskcc.org/.shortcut-targets-by-id/1L5uwxnNO44Pd7GBMBhMGO9vvsi7uifYZ/White Lab Gsuite Main Drive/Lab members/Miranda Hunter_GSuite/Miranda_R_new/bulk_rnaseq_datasets/confined_A375/' 
setwd(wd)

groups <- c("NC1", "NC2", "NC3", "C1", "C2", "C3") 

# import counts tables
for (group in groups) {
  path <- paste0(wd, group, "_ReadsPerGene.out.tab") # create a path saying where each counts matrix file is stored
  data <- read.delim(file = path, skip = 3) # read in the counts matrix file
  data[3:4] <- NULL # the last two columns of the matrix don't have useful info so just remove them
  colnames(data)[1:2] <- c("Ensembl", group) # rename columns to have better names
  assign(group,data) # store counts matrix for each group as the group name
  rm(data)
}

# create counts matrix containing all samples
counts_matrix <- merge(x = eval(parse(text = groups[1])),
                       y = eval(parse(text = groups[2])),
                       by = "Ensembl")

for (ii in 3:length(groups)) {
  group <- as.character(groups[ii])
  counts_matrix <- merge(x = counts_matrix,
                         y = eval(parse(text = group)),
                         by = "Ensembl")
}

counts_matrix <- counts_matrix %>% column_to_rownames(var = "Ensembl")

rm('C1', 'C2', 'C3', 'NC1', 'NC2', 'NC3')
```


Create DESeq2 dataset.
``` {r}
# create dataframe with your groups and the number of each replicate.
col_data <- data.frame(condition = c("NC", "NC", "NC", "C", "C", "C"), replicate = rep(c("1", "2", "3"), 2))
rownames(col_data) <- colnames(counts_matrix)
col_data$condition <- factor(col_data$condition)
col_data$replicate <- factor(col_data$replicate)
col_data$condition <- relevel(col_data$condition, ref = "NC")

deseq_matrix <- DESeqDataSetFromMatrix(countData = counts_matrix,
                                       colData = col_data,
                                       design = ~condition)
deseq_matrix
```

Pre filter low count genes
``` {r}
keep <- rowSums(counts(deseq_matrix)) >= 1 # remove any gene with 1 or fewer counts 
deseq_matrix <- deseq_matrix[keep, ]
```

Run DESeq2.
``` {r}
deseq_matrix <- DESeq(deseq_matrix, betaPrior = T)
resultsNames(deseq_matrix)
```

Make some comparisons across the groups.
``` {r}
# for contrast: whatever the fold change is relative to goes last!!!!
dp_res <- results(deseq_matrix, contrast = c("condition", "C", "NC"))

summary(dp_res)
```

Normalization and PCA.
``` {r}
deseq_matrix_norm <- vst(deseq_matrix, blind = F)

# PCA
pcaData <- plotPCA(object = deseq_matrix_norm,
                   intgroup = "condition",
                   returnData = F)
percentVar <- round(100 * attr(pcaData, "percentVar")) # calculate percent variance for the different PCs

pca_data <- plotPCA(object = deseq_matrix_norm,
                    intgroup = 'condition',
                    returnData = T)

ggplot(pca_data, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 4) +
  theme_classic() +
  xlab('PC1 (64% variance)') +
  ylab('PC2 (17% variance)') +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_line(size = 0.75),
        axis.title = element_text(size = 16, color = 'black'))
```


To get PC1 and PC2 genes: repeat PCA manually
``` {r}
vst_norm_matrix <- assay(deseq_matrix_norm) # extract normalized counts matrix from deseq_matrix_norm calculated above

pca_data_for_extracting_genes <- prcomp(t(vst_norm_matrix)) # re-run PCA to extract genes

# extract data with the contribution of each gene to the PCA calculation:
pca_gene_info <- pca_data_for_extracting_genes$rotation %>%
  data.frame() %>%
  rownames_to_column(var = "gene")

n_genes <- 5 # select number of genes you want to look at

# PC1:
PC1_high <- pca_gene_info %>% arrange(-PC1) %>% head(., n_genes) %>% pull(gene) # pull out the PC1 high genes
PC1_low <- pca_gene_info %>% arrange(-PC1) %>% tail(., n_genes) %>% pull(gene) # pull out the PC1 low genes

# PC2:
PC2_high <- pca_gene_info %>% arrange(-PC2) %>% head(., n_genes) %>% pull(gene) # pull out the PC2 high genes
PC2_low <- pca_gene_info %>% arrange(-PC2) %>% tail(., n_genes) %>% pull(gene) # pull out the PC2 low genes

# print results
message('PC1 high genes:')
message(paste(PC1_high, sep = " ", collapse = ", "))
message('PC1 low genes:')
message(paste(PC1_low, sep = " ", collapse = ", "))
message('PC2 high genes:')
message(paste(PC2_high, sep = " ", collapse = ", "))
message('PC2 low genes:')
message(paste(PC2_low, sep = " ", collapse = ", "))

```

Volcano plot:
``` {r}
for_volcano <- dp_res_df_named_confined

for_volcano$up_or_down <- case_when(
  for_volcano$log2FoldChange > 0 ~ 'up',
  TRUE ~ 'down'
)

genes_to_label <- for_volcano %>% filter(log2FoldChange > 0.5 & -log10(padj) > 11 | log2FoldChange < -0.5 & -log10(padj) > 3)


n_down <- for_volcano[!is.na(for_volcano$padj),] %>% filter(log2FoldChange < 0) %>% nrow()
n_up <- for_volcano[!is.na(for_volcano$padj),] %>% filter(log2FoldChange > 0) %>% nrow()

ggplot(for_volcano, aes(x = log2FoldChange, y = -log10(padj), color = up_or_down)) +
  #geom_hline(yintercept = -log10(0.05), linetype = 'dashed') +
  geom_point() +
  scale_color_manual(values = c('#3CBCCE', '#CE4E3C'),
                     labels = c(paste('downregulated genes\nn =', n_down), 
                                paste('upregulated genes\nn =', n_up))) +
  #scale_color_discrete() +
  #theme_minimal() +
  ylab('-log10 adjusted p-value') +
  xlab('log2 fold change') +
  theme(axis.text = element_text(size = 14, color = 'black'),
        axis.title = element_text(size = 16, color = 'black'),
        legend.title = element_blank(),
        legend.text = element_text(size = 12, color = 'black')) +
  geom_text_repel(data = genes_to_label, aes(label = hgnc_symbol),
                  box.padding = unit(0.5, 'lines'),
                  point.padding = unit(0.75, 'lines'),
                  max.overlaps = 10,
                  min.segment.length = 0,
                  color = 'black',
                  #seed = 101,
                  show.legend = F) 


```


Highlight genes w/sig p-value
``` {r}
for_volcano <- dp_res_df_named_confined

for_volcano$up_or_down <- case_when(
  for_volcano$log2FoldChange > 0 & for_volcano$padj < 0.05 ~ 'up',
  for_volcano$log2FoldChange < 0 & for_volcano$padj < 0.05 ~ 'down',
  TRUE ~ 'neither'
)

genes_to_label <- for_volcano %>% filter(log2FoldChange > 0.5 & -log10(padj) > 11 | log2FoldChange < -0.5 & -log10(padj) > 3)


n_down <- for_volcano[!is.na(for_volcano$padj),] %>% filter(log2FoldChange < 0 & padj < 0.05) %>% nrow()
n_up <- for_volcano[!is.na(for_volcano$padj),] %>% filter(log2FoldChange > 0 & padj < 0.05) %>% nrow()

ggplot(for_volcano, aes(x = log2FoldChange, y = -log10(padj), color = up_or_down)) +
  geom_hline(yintercept = -log10(0.05), linetype = 'dashed') +
  geom_point() +
  scale_color_manual(values = c('#3CBCCE',  'grey', '#CE4E3C'),
                     labels = c(paste('downregulated genes\nn =', n_down), 
                                paste('upregulated genes\nn =', n_up),
                                'neither')) +
  #scale_color_discrete() +
  #theme_minimal() +
  ylab('-log10 adjusted p-value') +
  xlab('log2 fold change') +
  theme(axis.text = element_text(size = 14, color = 'black'),
        axis.title = element_text(size = 16, color = 'black'),
        legend.title = element_blank(),
        legend.text = element_text(size = 12, color = 'black')) +
  geom_text_repel(data = genes_to_label, aes(label = hgnc_symbol),
                  box.padding = unit(0.5, 'lines'),
                  point.padding = unit(0.75, 'lines'),
                  max.overlaps = 10,
                  min.segment.length = 0,
                  color = 'black',
                  #seed = 101,
                  show.legend = F) 


```


## Add human gene names to counts table

Load gene list:
``` {r}
dp_res_df <- data.frame(dp_res) %>% rownames_to_column(var = "Ensembl")

```

Load conversion table:
``` {r}
library(biomaRt)
ensembl <- useEnsembl(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", mirror = "useast")
```


Extract gene symbols and ENSEMBL IDs:
``` {r}
data <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", 'description'),
              filters = "ensembl_gene_id",
              values = dp_res_df$Ensembl,
              mart = ensembl)
```

Add to DESeq2 table:
``` {r}
dp_res_df_named <- merge(x = dp_res_df,
                         y = data,
                         by.x = "Ensembl",
                         by.y = "ensembl_gene_id") %>% dplyr::select(Ensembl, hgnc_symbol, description, everything()) %>% arrange(-log2FoldChange)
```


Also add to counts matrix and normalized matrix:
``` {r}
counts_matrix <- counts_matrix %>% rownames_to_column(var = "Ensembl")
counts_matrix_named <- merge(x = counts_matrix,
                             y = data,
                             by.x = "Ensembl",
                             by.y = "ensembl_gene_id") %>% dplyr::select(Ensembl, hgnc_symbol, everything())

vst_norm_matrix <- data.frame(vst_norm_matrix) %>% rownames_to_column(var = "Ensembl")
vst_norm_matrix_named <- merge(x = vst_norm_matrix,
                         y = data,
                         by.x = "Ensembl",
                         by.y = "ensembl_gene_id") %>% dplyr::select(Ensembl, hgnc_symbol, everything())

```


Add ENSEMBL ID to hgnc_genes column in cases where there isn't a gene name. 
``` {r}
dp_res_df_named$hgnc_symbol <- case_when(
  dp_res_df_named$hgnc_symbol == "" ~ dp_res_df_named$Ensembl,
  TRUE ~ dp_res_df_named$hgnc_symbol
)

vst_norm_matrix_named$hgnc_symbol <- case_when(
  vst_norm_matrix_named$hgnc_symbol == "" ~ vst_norm_matrix_named$Ensembl,
  TRUE ~ vst_norm_matrix_named$hgnc_symbol
)

counts_matrix_named$hgnc_symbol <- case_when(
  counts_matrix_named$hgnc_symbol == "" ~ counts_matrix_named$Ensembl,
  TRUE ~ counts_matrix_named$hgnc_symbol
)
```


QC: plot number of counts per group.
``` {r}
total_UMIs <- colSums(counts_matrix[,-1]) %>% data.frame() %>% rownames_to_column(var = "group_name")
colnames(total_UMIs)[2] <- 'total_UMIs'
total_UMIs$group <- c("NC", "NC", "NC", "C", "C", "C")
total_UMIs$group <- factor(total_UMIs$group, levels = c("NC", "C"))

ggplot(total_UMIs, aes(x = group, y = total_UMIs, fill = group)) +
  geom_violin(scale = "width") +
  geom_point(size = 2.5) +
  ggtitle("total # of UMIs detected per sample") +
  ylab('# UMIs') +
  theme(axis.text.y = element_text(size = 16, color = "black"),
        axis.text.x = element_text(size = 18, color = "black"),
        axis.title.y = element_text(size = 18, color = "black"),
        axis.title.x = element_blank(),
        plot.title = element_text(size = 18, color = "black", face = "bold", hjust = 0.5),
        legend.position = "none") +
  coord_cartesian(ylim = c(0,60000000))

# also plot per sample
ggplot(total_UMIs, aes(x = group_name, y = total_UMIs, fill = group)) +
  geom_bar(stat = 'identity') +
  scale_fill_manual(values = c('#44A1F8', '#8B2152')) +
  theme(axis.text = element_text(size = 16, color = 'black'),
        axis.title = element_text(size = 18, color = 'black', face = 'bold'),
        axis.title.x = element_blank(),
        legend.position = 'none') +
  ylab('# of UMIs per sample')

```



## Start here if loading in previous results.
``` {r}
load('/Users/hunterm/Library/CloudStorage/GoogleDrive-hunterm@mskcc.org/.shortcut-targets-by-id/1L5uwxnNO44Pd7GBMBhMGO9vvsi7uifYZ/White Lab Gsuite Main Drive/Lab members/Miranda Hunter_GSuite/Miranda_R_new/bulk_rnaseq_datasets/confined_A375/counts_matrix_named_confined.R')
load('/Users/hunterm/Library/CloudStorage/GoogleDrive-hunterm@mskcc.org/.shortcut-targets-by-id/1L5uwxnNO44Pd7GBMBhMGO9vvsi7uifYZ/White Lab Gsuite Main Drive/Lab members/Miranda Hunter_GSuite/Miranda_R_new/bulk_rnaseq_datasets/confined_A375/dp_res_df_named_confined.R')
load('/Users/hunterm/Library/CloudStorage/GoogleDrive-hunterm@mskcc.org/.shortcut-targets-by-id/1L5uwxnNO44Pd7GBMBhMGO9vvsi7uifYZ/White Lab Gsuite Main Drive/Lab members/Miranda Hunter_GSuite/Miranda_R_new/bulk_rnaseq_datasets/confined_A375/vst_norm_matrix_named_confined.R')
```


Run GSEA: NO CUTOFF
Load pathways.
``` {r}
human.genes <- msigdbr(species = "Homo sapiens")
pathways.hallmark <- filter(human.genes, gs_cat == "H") %>% split(x = .$gene_symbol, f = .$gs_name) # Hallmark
pathways.GOBP <- filter(human.genes, gs_subcat == "GO:BP") %>% split(x = .$gene_symbol, f = .$gs_name) # GO BP
#pathways.GOCC <- filter(human.genes, gs_subcat == "GO:CC") %>% split(x = .$gene_symbol, f = .$gs_name) # GO CC
```


Make genelist for GSEA (same information as DESeq2 results above, just in a different format)
``` {r}
gsea_genelist <- dp_res_df_named_confined %>% dplyr::select(hgnc_symbol, log2FoldChange) %>% deframe()
```

GSEA with GO biological processes pathway set.
``` {r}
set.seed(100) # use the same set of random numbers every time for reproducibility

# GO biological pathways
gsea_gobp <- fgseaSimple(pathways = pathways.GOBP,
                           stats = gsea_genelist,
                         nperm = 10000) %>%
  mutate(log10pval = -log10(pval)) %>%
  dplyr::select(pathway, pval, log10pval, everything()) %>%
  arrange(desc(NES)) 
  
gsea_gobp
```

And with only significant genes:
``` {r}
dp_res_df_named_confined_sig <- dp_res_df_named_confined %>% filter(padj < 0.05)

gsea_gobp_sig <- fgseaSimple(pathways = pathways.GOBP,
                             stats = dp_res_df_named_confined_sig %>% dplyr::select(hgnc_symbol, log2FoldChange) %>% deframe(),
                             nperm = 10000) %>%
  mutate(log10pval = -log10(pval)) %>%
  dplyr::select(pathway, pval, log10pval, everything()) %>%
  arrange(desc(NES)) 

gsea_gobp_sig

```

Hallmark:
``` {r}
set.seed(100) # use the same set of random numbers every time for reproducibility

# GO biological pathways
gsea_hallmark <- fgseaSimple(pathways = pathways.hallmark,
                           stats = gsea_genelist,
                         nperm = 10000) %>%
  mutate(log10pval = -log10(pval)) %>%
  dplyr::select(pathway, pval, log10pval, everything()) %>%
  arrange(desc(NES)) 
  
gsea_hallmark

```

GO cellular component:
``` {r}
set.seed(100) # use the same set of random numbers every time for reproducibility

# GO biological pathways
gsea_gocc <- fgseaSimple(pathways = pathways.GOCC,
                           stats = gsea_genelist,
                         nperm = 1000) %>%
  mutate(log10pval = -log10(pval)) %>%
  dplyr::select(pathway, pval, log10pval, everything()) %>%
  arrange(desc(NES)) 
  
gsea_gocc
```

Run GSEA: 0.25 cutoff
Make genelist for GSEA (same information as DESeq2 results above, just in a different format)
``` {r}
cutoff <- 0.25
gsea_genelist_filt <- gsea_genelist[gsea_genelist > cutoff | gsea_genelist < -cutoff]

```

GSEA with GO biological processes pathway set.
``` {r}
set.seed(101) # use the same set of random numbers every time for reproducibility

# GO biological pathways
gsea_gobp_filt <- fgseaSimple(pathways = pathways.GOBP,
                           stats = gsea_genelist_filt,
                           nperm = 10000) %>%
  mutate(log10pval = -log10(pval)) %>%
  dplyr::select(pathway, pval, log10pval, everything()) %>%
  arrange(desc(NES)) 

gsea_gobp_filt
```


Plot top pathways.
``` {r}
go_term_bar(gsea_gobp_filt)
```


Plot top and bottom pathways.
``` {r}
gsea_plot <- rbind(head(gsea_gobp_filt,10),
                   tail(gsea_gobp_filt,10))
gsea_plot$pathway <- gsea_plot$pathway %>% gsub(pattern = "GOBP_", 
            replacement = "", x = ., ignore.case = F) %>% gsub(pattern = "_", 
            replacement = " ", x = .) %>% gsub(pattern = "plus", 
            replacement = "+", x = .) %>% tolower()
gsea_plot$pathway <- factor(gsea_plot$pathway, levels = rev(gsea_plot$pathway))

ggplot(gsea_plot, aes(x = NES, y = pathway, fill = NES)) +
  geom_bar(stat = 'identity') +
  xlab('normalized enrichment score') +
  scale_fill_gradientn(colors = rev(brewer.rdylbu(n = 100))) +
  theme(axis.text = element_text(size = 14, color = 'black'),
        axis.title.y = element_blank(),
        axis.title.x = element_text(size = 14, color = 'black'))

```



Save genelist for HOMER - just need ENS IDs. 
``` {r}
homer_list <- dp_res_df %>% filter(log2FoldChange > 0.25 | log2FoldChange < -0.25) %>% dplyr::select(Ensembl)
write.table(homer_list, file = 'confined_genelist_cutoff_025.txt', quote = F, row.names = F, col.names = F)

homer_list_up <- dp_res_df %>% filter(log2FoldChange > 0.25) %>% dplyr::select(Ensembl)
homer_list_down <- dp_res_df %>% filter(log2FoldChange < -0.25) %>% dplyr::select(Ensembl)

write.table(homer_list_up, file = 'confined_genelist_up_cutoff_025.txt', quote = F, row.names = F, col.names = F)
write.table(homer_list_down, file = 'confined_genelist_down_cutoff_025.txt', quote = F, row.names = F, col.names = F)

```


Plot top DE genes:
``` {r}
expr_plot_data <- rbind(head(dp_res_df_named_confined,10), tail(dp_res_df_named_confined,10))
expr_plot_data$hgnc_symbol <- factor(expr_plot_data$hgnc_symbol, levels = rev(expr_plot_data$hgnc_symbol))

expr_plot_data$change <- case_when(
  expr_plot_data$log2FoldChange > 0 ~ 'up',
  expr_plot_data$log2FoldChange < 0 ~ 'down'
)

ggplot(expr_plot_data, aes(y = hgnc_symbol, x = log2FoldChange, fill = log2FoldChange)) +
  geom_bar(stat = 'identity') +
  scale_fill_gradientn(colors = rev(brewer.rdbu(n = 20))) +
  theme(axis.text = element_text(color = 'black', size = 14),
        axis.title = element_text(color = 'black', size = 16),
        axis.title.y = element_blank(),
        legend.text = element_text(size = 10, color = 'black'),
        legend.title = element_text(size = 12, color = 'black')) +
  labs(fill = 'log2FC')
  
```





