---
title: "Reanalysis of Jerby human data"
author: "Miranda Hunter"
date: "10/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries:
``` {r}
library(Seurat)
library(tidyverse)
library(scales)
library(pals)
library(cowplot)
library(fgsea)
library(msigdbr)
library(data.table)
library(mvhfunctions)
library(R.utils)
```

## Import and process data.

Import data:
``` {r}
counts <- fread('/Users/hunterm/Library/CloudStorage/GoogleDrive-hunterm@mskcc.org/.shortcut-targets-by-id/1L5uwxnNO44Pd7GBMBhMGO9vvsi7uifYZ/White Lab Gsuite Main Drive/Lab members/Miranda Hunter_GSuite/Miranda_R_new/singlecell_datasets/Jerby_data/GSE115978_counts.csv.gz', stringsAsFactors = F) %>% column_to_rownames(var = "V1")

meta <- fread('/Users/hunterm/Library/CloudStorage/GoogleDrive-hunterm@mskcc.org/.shortcut-targets-by-id/1L5uwxnNO44Pd7GBMBhMGO9vvsi7uifYZ/White Lab Gsuite Main Drive/Lab members/Miranda Hunter_GSuite/Miranda_R_new/singlecell_datasets/Jerby_data/GSE115978_cell.annotations.csv.gz', stringsAsFactors = F) %>% column_to_rownames(var = "cells")
```

Rename "Mal" to tumor.
``` {r}
meta$cell.types <- gsub(pattern = "Mal", replacement = "Tumor", x = meta$cell.types)
```

Create Seurat object.
``` {r}
jerby <- CreateSeuratObject(counts = counts,
                            project = "Jerby-Arnon_Seurat",
                            meta.data = meta)
jerby
```

Processing of data:
``` {r, message = F, warning = F}
jerby <- SCTransform(jerby, verbose = F)
jerby <- FindVariableFeatures(jerby, verbose = F)
jerby <- RunPCA(jerby, verbose = F)
ElbowPlot(jerby)
```

``` {r}
jerby <- FindNeighbors(jerby, dims = 1:10, verbose = F)
jerby <- FindClusters(jerby, verbose = F)
jerby <- RunUMAP(jerby, verbose = F, dims = 1:10)
nice_dim_plot(jerby, group_by = "cell.types", label = F)
```


Look at suspected "interface" cluster.
``` {r}
cluster_res <- seq(0.3, 1.2, 0.1)

for (res in cluster_res) {
  jerby <- FindClusters(jerby, resolution = res, verbose = F)
}

plotlist <- NULL
for (res in cluster_res) {
  index <- which(cluster_res == res)
  plotlist[[index]] <- nice_dim_plot(jerby, 
                                     group_by = paste0('SCT_snn_res.', res), 
                                     label = T) + NoLegend() + ggtitle(paste(res, 'resolution'))
}
plot_grid(plotlist = plotlist, nrow = 2)
```


``` {r}
DimPlot(jerby, group.by = c('cell.types', 'SCT_snn_res.0.8'), label = T, label.box = T)
```

Cluster 9 looks interesting - looks like potential interface cluster. Add to metadata and calculate markers.
``` {r}
jerby$cell_types_new <- case_when(
  jerby$SCT_snn_res.0.8 == 9 ~ 'interface',
  jerby$SCT_snn_res.0.8 == 16 ~ 'CAFs',
  jerby$SCT_snn_res.0.8 %in% c(5,6,10,17,8,11,12) ~ 'tumor',
  jerby$SCT_snn_res.0.8 == 15 ~ 'endo',
  TRUE ~ 'immune'
)

DimPlot(jerby, group.by = 'cell_types_new')

```

Subset out just tumor and interface clusters.
``` {r}
Idents(jerby) <- "cell_types_new"

jerby_TI <- subset(jerby, idents = c('tumor', 'interface'))
```


## Calculate interface markers.
``` {r}
Idents(jerby) <- "cell_types_new"
interface_markers_relative_to_tumor <- FindMarkers(jerby,
                                                   ident.1 = 'interface',
                                                   ident.2 = 'tumor',
                                                   assay = 'SCT') %>%
  rownames_to_column(var = "gene") %>%
  arrange(-avg_log2FC)

interface_markers_all <- FindMarkers(jerby,
                                     ident.1 = 'interface',
                                     assay = 'SCT') %>%
  rownames_to_column(var = "gene") %>%
  arrange(-avg_log2FC)

# save so don't have to rerun this again
setwd('/Users/hunterm/Library/CloudStorage/GoogleDrive-hunterm@mskcc.org/.shortcut-targets-by-id/1L5uwxnNO44Pd7GBMBhMGO9vvsi7uifYZ/White Lab Gsuite Main Drive/Lab members/Miranda Hunter_GSuite/Miranda_R_new/singlecell_datasets/Jerby_data/')
save(interface_markers_relative_to_tumor, file = 'interface_markers_relative_to_tumor.R')
save(interface_markers_all, file = 'interface_markers_all.R')
```


## Plot top 20 genes upregulated in interface.
``` {r}
interface_markers_plot <- head(interface_markers_relative_to_tumor,20) 
interface_markers_plot$gene <- factor(interface_markers_plot$gene, levels = interface_markers_plot$gene)

interface_markers_plot$neuronal <- case_when(
  interface_markers_plot$gene %in% c('NNAT', 'NEUROD1', 'NEFM', 'PPP1R17', 'SOX11', 'NEFL', 'CHRNA1') ~ 'neuronal',
  TRUE ~ 'not neuronal'
)

ggplot(interface_markers_plot, aes(x = gene, y = avg_log2FC, fill = neuronal)) +
  geom_bar(stat = 'identity') +
  scale_fill_manual(values = c('#4D2F67', '#C8C8C8')) +
  theme_minimal() +
  theme(axis.text = element_text(size = 14, color = 'black'),
        axis.title = element_text(size = 16, color = 'black'),
        axis.text.x = element_text(angle = -45, hjust = 0, vjust = 0.5),
        axis.title.y = element_blank()) +
  xlab('log2FC in interface') 
```


## Clean up cell type annotations (original cell types)
``` {r}

jerby$cell.types <- gsub(pattern = 'Endo.', replacement = 'Endothelial', x = jerby[[]]$cell.types) %>%
  gsub(pattern = 'T.CD4', replacement = 'CD4 T cell', x = .) %>%
  gsub(pattern = 'T.CD8', replacement = 'CD8 T cell', x = .) %>%
  gsub(pattern = 'B.cell', replacement = "B cell", x = .) %>%
  gsub(pattern = 'NK', replacement = 'NK cell', x = .) %>%
  gsub(pattern = 'T.cell', replacement = "T cell", x = .)


# also add another cell type annotation calling interface "Tumor (HMGB2-hi)"
jerby$cell_types_new_HMG <- gsub(pattern = 'interface', replacement = 'tumor (HMGB2-high)', x = jerby$cell_types_new)

nice_dim_plot(jerby, group_by = 'cell.types', label = F)
nice_dim_plot(jerby, group_by = 'cell_types_new_HMG', label = F)
```


## GSEA: interface cluster
``` {r}
human.genes <- msigdbr(species = "Homo sapiens")
pathways.GOBP <- filter(human.genes, gs_subcat == "GO:BP") %>% split(x = .$gene_symbol, f = .$gs_name) # GO BP
```


``` {r}
set.seed(100)

# remove ribosomal genes
interface_tumor_genelist <- interface_markers_relative_to_tumor %>% 
  dplyr::select(gene, avg_log2FC) %>% 
  .[!startsWith(.$gene, c('RPS', 'RPL')),]

interface_tumor_gsea <- fgseaMultilevel(pathways = pathways.GOBP,
                                        stats = deframe(interface_tumor_genelist)) %>%
  arrange(-NES)
interface_tumor_gsea
```


## GSEA barcode plots:
``` {r}
limma::barcodeplot(statistics = deframe(interface_tumor_genelist), 
                   index = pathways.GOBP[['GOBP_CELL_FATE_SPECIFICATION']],
                   quantiles = c(0,0),
                   main = 'GOBP_CELL_FATE_SPECIFICATION')

limma::barcodeplot(statistics = deframe(interface_tumor_genelist), 
                   index = pathways.GOBP[['GOBP_REGULATION_OF_NEURON_DIFFERENTIATION']],
                   quantiles = c(0,0),
                   main = 'GOBP_REGULATION_OF_NEURON_DIFFERENTIATION')
```


## Plot expression of melanocyte genes.
``` {r}
plots <- FeaturePlot(jerby_TI, 
            features = c('MITF', 'TYRP1', 'PMEL',  'SOX9', 'SOX10', 'SOX11'),
            order = T,
            pt.size = 0.4,
            combine = F)
plots <- lapply(plots, function(x) {
  x + scale_color_gradientn(colors = viridis(n = 100))
})
plot_grid(plotlist = plots, align = 'hv', axis = 'lrbt', nrow = 2)
```


## Map interface genes from Nat Comms paper onto this dataset.
``` {r}
load('/Users/hunterm/Library/CloudStorage/GoogleDrive-hunterm@mskcc.org/.shortcut-targets-by-id/1L5uwxnNO44Pd7GBMBhMGO9vvsi7uifYZ/White Lab Gsuite Main Drive/Lab members/Miranda Hunter_GSuite/Miranda_R_new/seuratobjs/EF_integrated_allgenes_new.R')

Idents(EF.filt) <- "integrated_cell_type"
SC_interface_genes <- FindMarkers(EF.filt,
                                  ident.1 = "interface",
                                  assay = "SCT") %>%
  rownames_to_column(var = "gene") %>%
  filter(p_val_adj <= 0.05) %>%
  arrange(-avg_log2FC)

jerby_TI <- AddModuleScore(jerby_TI,
                        features = list(SC_interface_genes %>% filter(avg_log2FC >= 0.5) %>% pull(gene) %>% convert_to_human_list_MVH() %>% intersect(rownames(jerby_TI), .)),
                        name = "SC_interface")

nice_feature_plot(jerby_TI, features = "SC_interface1", pt.size = 0.5)
```

Scale the interface expression score and add to metadata.
``` {r}
jerby_TI$SC_interface_scaled <- rescale(jerby_TI$SC_interface1, to = c(0,1))

nice_feature_plot(jerby_TI, features = 'SC_interface_scaled', pt.size = 1.5) + scale_color_gradientn(colors = inferno(n = 100))
```



## Score cells with Tsoi gene sets and classify into states.
``` {r}
tsoi_genelists <- readxl::read_excel('/Users/hunterm/Library/CloudStorage/GoogleDrive-hunterm@mskcc.org/.shortcut-targets-by-id/1L5uwxnNO44Pd7GBMBhMGO9vvsi7uifYZ/White Lab Gsuite Main Drive/Lab members/Miranda Hunter_GSuite/Miranda_R_new/Tsoi_2018_genesets.xlsx')
```

Scoring:
``` {r}
colnames(tsoi_genelists) <- tsoi_genelists[1,]
tsoi_genelists <- tsoi_genelists[2:nrow(tsoi_genelists),]

sigs <- unique(tsoi_genelists$Signature)

for (geneset in sigs) {
  
  genes <- tsoi_genelists %>% filter(Signature == geneset) %>% pull(Gene)
  jerby_TI <- AddModuleScore(jerby_TI,
                             features = list(intersect(genes, rownames(jerby))),
                             name = geneset)
}
```

Scale scores between 0 and 1 and add to metadata.
``` {r}
tsoi_sig_scores <- jerby_TI[[]] %>% 
  #rownames_to_column(var = 'cell') %>%
  dplyr::select(all_of(paste0(sigs,'1'))) 

tsoi_sig_scores_scaled <- apply(tsoi_sig_scores, 2, rescale, to = c(0,1)) %>% data.frame()

jerby_TI <- AddMetaData(jerby_TI,
                        metadata = tsoi_sig_scores_scaled,
                        col.name = paste0(sigs, '_scaled'))

```

Plot scores for each state in interface cells only.
``` {r}
scores_plot_scaled <- jerby_TI[[]] %>% 
  dplyr::select(all_of(c('cell_types_new', paste0(sigs, '_scaled')))) %>%
  melt()

scores_plot_scaled$variable <- gsub('_scaled', '', scores_plot_scaled$variable)
scores_plot_scaled$variable <- factor(scores_plot_scaled$variable, levels = rev(sigs))

ggplot(scores_plot_scaled %>% dplyr::filter(cell_types_new == 'interface'), aes(y = value, x = variable, fill = variable)) +
  geom_violin(scale = 'width', aes(color = variable)) +
  geom_jitter(size = 0.00001) +
  ggtitle('Tsoi cell state module scores - interface') +
  scale_fill_manual(values = tol(n = length(sigs))) +
  scale_color_manual(values = tol(n = length(sigs))) +
  ylab('Tsoi cell state module score') +
  theme(axis.text = element_text(size = 12, color = 'black'),
        axis.text.x = element_text(angle = -45, hjust = 0, vjust = 0),
        axis.title.x = element_blank(),
        axis.title = element_text(size = 14, color = 'black'),
        legend.position = 'none',
        plot.title = element_text(size = 15),
        plot.margin = margin(r = 75, l = 20, t = 20, b = 20))
```


Classify cells based on which state they have the highest score for.
``` {r}
scores_all <- jerby_TI[[]] %>% dplyr::select(Undifferentiated_scaled, `Neural crest-like_scaled`, Transitory_scaled, Melanocytic_scaled)
scores_all$highest_score <- colnames(scores_all)[apply(scores_all, 1, which.max)]

jerby_TI <- AddMetaData(jerby_TI,
                        metadata = scores_all %>% dplyr::select(highest_score),
                        col.name = 'highest_module')
```

Plot:
``` {r}
nice_dim_plot(jerby_TI, group_by = 'highest_module', label = F)
```

## Plot HMGB2 expression across cell types.
``` {r}
jerby$cell_types_new <- factor(jerby$cell_types_new, levels = c('tumor', 'interface', 'CAFs', 'endo', 'immune'))

p1 <- nice_feature_plot(jerby, features = 'HMGB2', order = F, pt.size = 0.75) + scale_color_gradientn(colors = viridis(n = 100))
p2 <- VlnPlot(jerby, features = 'HMGB2', group.by = 'cell_types_new', pt.size = 0, cols = tol(5)) + NoLegend()
plot_grid(p1,p2, nrow = 1, align = 'hv', axis = 'lrbt')
```